### 2023-11-08 manuscript/manuscript.Rmd
rewrite of section 3.2 to include the chemical standards assignment results in the standards mix

```{r assignment-success}
tar_load(standards_compound_PIPs)
tar_load(spiked_urine_feature_matches)
# tar_load(spiked_urine_unassigned)
# tar_load(spiked_urine_assigned)
# tar_load(spiked_urine_incorrect_assignments)

brachy_shared_MFs <- tar_read(brachy_FIE_HRMS_results_molecular_formula_assignment) %>% 
  assignments::assignments() %>% 
  select(MF) %>% 
  inner_join(tar_read(brachy_RP_UHPLC_HRMS_results_molecular_formula_assignment) %>% 
               assignments::assignments() %>% 
               select(MF),
             by = 'MF') %>% 
  distinct() %>% 
  nrow()

brachy_shared_kegg <- intersect(
  tar_read(brachy_FIE_HRMS_KEGG_matches)$MF %>% 
    unique(),
  tar_read(brachy_RP_UHPLC_HRMS_KEGG_matches)$MF %>% 
    unique()
) %>% length()

positive_mode_adducts <- brachy_FIE_HRMS_parameters_molecular_formula_assignment %>% 
 assignments::adducts() %>% 
  .$p %>% 
  length()

negative_mode_adducts <- brachy_FIE_HRMS_parameters_molecular_formula_assignment %>% 
 assignments::adducts() %>% 
  .$n %>% 
  length()

spiked_urine_MF_PIPs <- standards_compound_PIPs %>% 
  select(MF,Adduct,Isotope) %>% 
  distinct()

spiked_urine_feature_matches_MF_n <- spiked_urine_feature_matches$MF %>% 
  unique() %>% length()

# percent_unassigned <- {nrow(spiked_urine_unassigned) / 
#     nrow(spiked_urine_feature_matches) * 100} %>% 
#   signif(digits = 3)

# spiked_urine_unassigned_MF_n <- spiked_urine_unassigned %>% 
#   anti_join(
#     spiked_urine_assigned %>% 
#       select(MF) %>% 
#       distinct(), 
#     by = 'MF') %>% 
#   .$MF %>% 
#   unique() %>% 
#   length()

# spiked_urine_correct_assignments_MF_n <- spiked_urine_correct_assignments %>%
#   .$MF %>% 
#   unique() %>% 
#   length()

# spike_urine_standards_MF_assignment_impossible_n <- spiked_urine_feature_matches_MF_n -
#   nrow(spike_urine_standards_MF_assignment_possible)

# percent_standards_MF_assigned <- {spiked_urine_correct_assignments_MF_n / 
#   nrow(spike_urine_standards_MF_assignment_possible) * 100} %>% 
#   signif(digits = 3)
```

An overview of the molecular formula assignments made in both the urine and *B. distachyon* FIE-HRMS matrices is provided in Table \@ref(tab:assignment-results), which shows that similar proportions of *m/z* features were assigned molecular formulas.
However, there was more than double the relative molecular formula assignment coverage for the C18-UHPLC-HRMS data of *B. distachyon* compared to the FIE-HRMS matrices.
The chromatographic separation of the C18-UHPLC-HRMS increases the likelihood that a number of correlated ionisation products will be present for a given compound due to a reduction in the effects of ion suppression [@gowda2014].
There is also a decreased chance of interference in the correlations between related features due to the presence of structural isomers and isobaric compounds.
A total of `r brachy_shared_MFs` molecular formulas were assigned in both the FIE-HRMS and C18-UHPLC-HRMS data of *B. distachyon*.

@@

```{r assignment-evaluations}
tar_load(assignment_outcomes_summary)

@@

@@

```{r compound-evaluations}
tar_load(standards_compound_outcomes)

standards_compound_outcomes %>%
  dplyr::mutate(
    outcome = outcome %>%
      stringr::str_replace_all(
        'm/z',
        '\\\\textit{m/z}'
      ) %>% 
      stringr::str_replace_all(
        '&',
        '\\\\&'
      )
  ) %>%
  dplyr::relocate(
    spiked_urine,
    .after = dplyr::everything()
  ) %>%
  dplyr::rename(
    ` ` = outcome,
    `Standards mix` = standards,
    `Spiked urine` = spiked_urine
  ) %>%
  kableExtra::kable(
    format = 'latex',
    booktabs = TRUE,
    escape = FALSE,
    linesep = '',
    align = c(
      rep('l',1),
      rep('r',2)
    ),
    caption = 'Molecular formula assignment results frequencies for the 64 unique molecular formulas of the chemical standards in the standards mix and spiked into human urine.'
  ) %>%
  kableExtra::kable_styling(
    latex_options = "HOLD_position"
  )
```

@@

Although there were proportionately more molecular formulas assigned to *m/z* features in the C18-UHPLC-HRMS compared to the FIE-HRMS *B. distachyon* matrix, there was a disproportionate number of these molecular formulas that matched to compounds in the KEGG database specific to *B. distachyon* and able to form the relevant ionisation products (Table \@ref(tab:assignment-results)).
However, only `r brachy_shared_kegg` of the molecular formulas matched to compounds in the KEGG database were found to be common between the two techniques.
The low number of assigned molecular formulas partly reflects the incompleteness of the KEGG database, especially with respect to the secondary metabolites that are prolific in C18-UHPLC-HRMS [@breitling2013].
The assignments for all the matrices along with the matched KEGG compound IDs are provided in Additional file 4.

To assess the assignments in the example urine matrix with respect to the spiked chemical standards, the predicted ionisation products were first identified by directly matching these to the observed *m/z*.
Based on the `r positive_mode_adducts + negative_mode_adducts` adducts and four isotopic rules specified, there were `r nrow(spiked_urine_MF_PIPs)` predicted ionisation product *m/z* for these chemical standards.
A total of `r nrow(spiked_urine_feature_matches)` of these ionisation products could be directly matched to the observed *m/z* features which included `r spiked_urine_feature_matches_MF_n` of the `r standards_compound_MFs` unique molecular formulas of the spiked chemical standards. 
These directly matched *m/z* features are provided in Additional file 5.
For `r #nrow(spiked_urine_unassigned)` (`r #percent_unassigned`%) of these matched *m/z* features, no molecular formula was assigned by the approach, which included `r #spiked_urine_unassigned_MF_n` of the molecular formulas of the spiked chemical standards.
However, `r #nrow(spiked_urine_assigned)` of the matched features were assigned a molecular formula with `r #nrow(spiked_urine_correct_assignments)` of these corresponding to one of the spiked compound ionisation products. This included `r #spiked_urine_correct_assignments_MF_n` of the spiked chemical standards molecular formulas (Table \@ref(tab:assignment-results)).
The *m/z* features assigned to molecular formulas corresponding to the spiked chemical standards are provided in Additional file 6.

The remaining `r #nrow(spiked_urine_incorrect_assignments)` matched *m/z* features were assigned alternative molecular formula ionisation products.
The *m/z* of the predicted ionisation features that it was not possible to assign molecular formulas were as a result of either not being correlated with other *m/z* features that corresponded to a correct adduct or isotope mass difference or whose subgraph components were not selected, with more plausible solutions instead selected based on the correlation network.

Accounting for the `r #spike_urine_standards_MF_assignment_impossible_n` chemical standards that were not assigned due to a lack of relevant correlated mass relationships, the algorithm was able to assign molecular formulas to features corresponding to `r #percent_standards_MF_assigned`% of the `r #nrow(spike_urine_standards_MF_assignment_possible)` possible chemical standards.
This highlights the difficulty in annotating these complex biological mixtures containing hundreds of compounds, which can create many thousands of possible ionisation products.
Interference in correlations due to the presence of isobaric compounds and MS instrument noise can weaken correlations between features, particularly in those with low abundance.

@@

The use of correlation analysis to identify related features requires a suitable number of observations to ensure that statistically significant associations can be detected. 
This is especially important in non-targeted metabolomics data sets that can contain thousands of *m/z* features and p-value adjustment is required to account for multiple testing. 
For example, to detect a significant lower limit of Pearson's correlation coefficient of 0.7 for a matrix containing `r sample_size_n_features` *m/z* features, an approximate minimum sample size of `r sample_size` observations would be required (p < 0.05, Bonferroni corrected) [@cohen2013]. 
Metabolomics fingerprinting and profiling experiments typically require high replication of 10-15 observations per class in order to discriminate between the sample classes [@enot2008]. 
Therefore, a sample set comparing just four classes brings this to within a practical range for applying this approach in many cases, such as the example *B. distachyon* ecotype comparison sample set provided here.
The assignments obtained from the *B. distachyon* example matrix also show the application of the assignment approach to sample sets that contain multiple underlying experimental classes.

>>>>

```{r assignment-evaluations}
tar_load(standards_parameters_molecular_formula_assignment)
tar_load(standards_assignment_outcomes_non_iso)
tar_load(spiked_urine_assignment_outcomes_non_iso)
tar_load(standards_compound_outcomes)

n_adducts <- standards_parameters_molecular_formula_assignment %>%
  assignments::adducts() %>% 
  purrr::map_dbl(length) %>% 
  sum()

n_features <- list(
  standards = standards_assignment_outcomes_non_iso,
  spiked_urine = spiked_urine_assignment_outcomes_non_iso
) %>% 
  purrr::map(
    ~.x %>% 
      dplyr::pull(
        Feature
      ) %>%
      unique() %>%
      length()
  )

standards_n_AI_relevant <- standards_assignment_outcomes_non_iso %>% 
  dplyr::filter(
    correlations == TRUE,
    relationships == TRUE
  ) %>% 
  dplyr::pull(MF) %>% 
  unique() %>% 
  length()

standards_n_AI_assigned <- standards_assignment_outcomes_non_iso %>% 
  dplyr::filter(
    matching_assignment == TRUE
  ) %>% 
  dplyr::pull(MF) %>% 
  unique() %>% 
  length()

standards_percent_AI_assigned <- {standards_n_AI_assigned / standards_n_AI_relevant * 100} %>% 
  signif(digits = 3)

standards_n_alternative <- standards_assignment_outcomes_non_iso %>% 
  dplyr::filter(
    dplyr::if_all(
      correlations:assigned,
      ~.x == TRUE
    ),
    matching_adduct == FALSE
  ) %>% 
  nrow()

compounds_percent_assigned <- standards_compound_outcomes %>% 
  dplyr::filter(
    outcome %in% c(
      "Matched to m/z features",
      "Assigned"
    )
  ) %>% 
  tidyr::gather(
    sample,n,
    -outcome
  ) %>% 
  tidyr::spread(
    outcome,n
  ) %>% 
  dplyr::group_by(sample) %>% 
  dplyr::summarise(
    percent = {Assigned / `Matched to m/z features` * 100} %>% 
      signif(digits = 3)
  ) %>% 
  dplyr::pull(percent) %>% 
  purrr::set_names(c(
    'spiked_urine',
    'standards'
  )) %>% 
  as.list()
```

To assess the performance of the molecular formula and graphical component selection approaches, the assignment outcomes were evaluated for the theoretical ionisation products calculated for the chemical standards that matched to *m/z* features in the standards mix and spiked urine. 
The standards mix allows for a direct evaluation of the selection approaches in a relatively simple, synthetic matrix.
Alternatively, the spiked urine provides an indication of how the selection approaches perform in a more realistic context where there are more *m/z* features present greater competition for assignment. 
To simplify interpretation, these evaluations were based on the assignment outcomes of the non-isotopic ionisation products during the adduct and isotopic assignment iterations.
This reduces the incidence of overlap between ionisation products, where multiple theoretical ionisation products match to a single *m/z* feature as well as a removes the inclusion of relationships involving transformations that complicates the calculation and interpretation of relevant relationships.
Based on the `r n_adducts` adduct rules specified, there were `r nrow(standards_assignment_outcomes_non_iso)` and `r nrow(spiked_urine_assignment_outcomes_non_iso)` theoretical ionisation products for these chemical standards that matched to `r n_features$standards` and `r n_features$spiked_urine` *m/z* in the standards mix and spiked urine respectively.
The theoretical ionisation products that directly matched *m/z* features, including the isotopic theoretical ionisation products, are provided in Additional file 5.
The frequencies of assignment outcomes for the theoretical ionisation products in these matrices are shown in Table \@ref(tab:ip-outcomes).

@@

```{r ip-outcomes}
tar_load(assignment_outcomes_summary)

@@

Of the `r standards_n_AI_relevant` chemical standards that had theoretical ionisation products with relevant relationships in the standards mix, the molecular formula and graphical component selection approaches were able to correctly assign molecular formulas to `r standards_n_AI_assigned` (`r standards_percent_AI_assigned`%) of these.
For the `r xfun::numbers_to_words(standards_n_alternative)` ionisation products for which an alternative adduct and molecular formula were assigned, two of these were assigned to alternative components with a higher $P_c$ score resulting from a higher component AIS score.
The other alternative component was selected due to higher $P_c$ score resulting from a higher component degree.
There was only a single instance where an ionisation product was incorrectly assigned as a result of the molecular formula selection approach.

The incidence of alternative adduct and molecular formula assignment is higher in the spiked urine (Table \@ref(tab:ip-outcomes)).
However, this is expected as there and are more *m/z* features present and greater numbers of background compounds and therefore greater chance for alternative graphical components to be present in the correlation network.
It does not necessarily reflect a poor performance of the graphical component selection approach, there is just a higher probability that these *m/z* could potentially be alternative compounds. 
Alternative assignment resulting from molecular formula selection still remains low.

The assignment results for the chemical standards in the standards mix and spiked urine are provided in Table \@ref(tab:compound-frequencies).
These show that, where compounds had *m/z* features matching to theoretical ionisation products, the approach was able to assign `r compounds_percent_assigned$standards`% and `r compounds_percent_assigned$spiked_urine`% of these chemical standards in the standards mix and spiked urine respectively. 
The *m/z* for which a chemical standard was assigned are provided in Additional file 6.

@@

```{r compound-frequencies}
standards_compound_outcomes %>%
  dplyr::mutate(
    outcome = outcome %>%
      stringr::str_replace_all(
        'm/z',
        '\\\\textit{m/z}'
      ) %>% 
      stringr::str_replace_all(
        '&',
        '\\\\&'
      )
  ) %>%
  dplyr::relocate(
    spiked_urine,
    .after = dplyr::everything()
  ) %>%
  dplyr::rename(
    ` ` = outcome,
    `Standards mix` = standards,
    `Spiked urine` = spiked_urine
  ) %>%
  kableExtra::kable(
    format = 'latex',
    booktabs = TRUE,
    escape = FALSE,
    linesep = '',
    align = c(
      rep('l',1),
      rep('r',2)
    ),
    caption = 'Molecular formula assignment results frequencies for the 64 unique molecular formulas of the chemical standards in the standards mix and spiked into human urine.'
  ) %>%
  kableExtra::kable_styling(
    latex_options = "HOLD_position"
  )
```

@@

An overview of the molecular formula assignments made across the example sample matrices is provided in Table S6 (Additional file 3).
All of the assignments for these example matrices along with any matched KEGG compound IDs are provided in Additional file 4.
The assignments obtained from the *B. distachyon* is provided as an example showing the intended real world application of the molecular formula assignment approach to a sample set that includes multiple underlying experimental classes in the correlation network.
A similar number of molecular formulas were assigned to the FIE-HRMS matrices for the spiked human urine and *B. distachyon* matrices as well as similar numbers of molecular formulas that were matched to KEGG compounds.

While not the intended primary application of the approach, the C18-UPHLC-HRMS data from *B. distachyon* had the highest percentage of *m/z* features assigned along with the greatest number of molecular formulas. 
The chromatographic separation of the C18-UHPLC-HRMS enables the correlation network to be considered within limited retention time windows and increases the likelihood that a number of correlated ionisation products will be present for a given compound.
Added to this, the LC-MS reduces the effects of ion suppression compare to FIE-HRMS and also decreases the chance of potential interference in the correlations between related features due to the presence of structural isomers and isobaric compounds [@gowda2014].

@@

The use of correlation analysis to identify related features requires a suitable number of observations to ensure that statistically significant associations can be detected. 
This is especially important in non-targeted metabolomics data sets that can contain thousands of *m/z* features and p-value adjustment is required to account for multiple testing. 
For example, to detect a significant lower limit of Pearson's correlation coefficient of 0.7 for a matrix containing `r sample_size_n_features` *m/z* features, an approximate minimum sample size of `r sample_size` observations would be required (p < 0.05, Bonferroni corrected) [@cohen2013]. 
Metabolomics fingerprinting and profiling experiments typically require high replication of 10-15 observations per class in order to discriminate between the sample classes [@enot2008]. 
Therefore, a sample set comparing just four classes brings this to within a practical range for applying this approach in many cases, such as the example *B. distachyon* ecotype comparison sample set provided here.
